Adaptive group-regularized logistic elastic net regression Supplementary material

### 2019-02-27 docs/response.Rnw

\documentclass{letter}

\newenvironment{itquote}
{\begin{quote}\itshape}
{\end{quote}}

\signature{Magnus M\"unch}
\begin{document}
  \begin{letter}{}
    \opening{Dear editor,}

    Thank you for the opportunity to resubmit our manuscript. We realise it has
    taken us an extroardinarily long time to respond, but given the extensive 
    nature of the revision, we hope that you will still consider this 
    resubmission.

    Below we formulate our response to your and the reviewers' comments
    point by point. We hope that these responses and the accompanying changes in
    the manuscript and supplementary material address the raised issues, such 
    that the manuscript is deemed appropriate for publication in Biostatistics.

    \textbf{Reply to editor}
    \begin{itquote}
      I would be willing to consider a revision of your work if you 
      made a substantial effort to rework the simulation examples and practical 
      case studies to demonstrate clearly unique cases where your method works 
      and others don't (as well as the reverse case!).
    \end{itquote} 
    The simulations were extended to five different scenarios. In three
    scenarios, we expect all methods to underperform. In one scenario, we expect
    \texttt{gren} to underperform compared to its competitors and in one 
    scenario we expect \texttt{gren} to outperform its competitors. We have 
    included 4/5 applications in the manuscript and supplementary material. In
    one application \texttt{gren} underperforms compared to the others. We argue
    as for why this happens and deduce certain scenarios in which \texttt{gren}
    should not be used.

    \begin{itquote}
      This comparison should include, not just the EN and Grouped 
      Lasso, but other methods that approach this same problem as well. There 
      would need to be clear empirical demonstration on real data of a unique 
      advantage of this approach for it to be a good fit at Biostatistics.
    \end{itquote}
    We included comparisons to the sparse group lasso (\texttt{SGL}) by 
    Simon et al. (2013), the group mimimax concave penalty
    (\texttt{grMCP}) by Breheny and Huang (2009), and the group exponential
    lasso (\texttt{gel}) by Breheny (2015). \texttt{gren} outperforms these in 
    a number of comparisons. Our approach is based on a different philosophy
    than the group lasso and its extensions. Group weights are adaptively 
    estimated from the data and not based on one common penalty, as is the case
    in the group lasso and its extensions.

    \textbf{Reply to reviewer 1}
    \begin{itquote}
      However, in my opinion the scientific contribution and novelty 
      of the methodology is just incremental. Elastic net itself is designed to 
      do the grouping. So, I am not sure why putting group penalty elastic net
      will be better than other group penalty methods like group lasso
    \end{itquote}
    The grouping property of elastic net refers to a theoretical advantage over
    the lasso in dealing with correlated features. It does not refer to the
    explicit inclusion of external information in the form of groups of 
    features and is therefore unrelated to the topic of the manuscript.

    \begin{itquote}
      Also Bayesian logistic regression can be formulated based on 
      Polya-Gamma latent variables (Polson et al.) and that need to be 
      incorporated in your modeling strategy.
    \end{itquote}
    The Polya-Gamma parametrization is actually an important part of our method.
    It is introduced in equation (6) in the Section 3.2 and extensively covered 
    in Section 4 of the supplementary material.

    \begin{itquote}
      Why no comparison between the proposed method and at least the 
      group lasso which assigns the group penalty.
    \end{itquote}
    We included comparisons to the group lasso and several of its extensions in
    the simulations and data applications.

    \begin{itquote}
      The model contains several tuning/prior parameters. Sensitivity 
      analysis need to be provided for that.
    \end{itquote}
    We show in one of our applications that randomly assigning the features
    to the groups does not lead to bias in the tuning parameter estimates, 
    suggesting that the method is robust (at least in this application).

    \begin{itquote}
      How much your model is scalable? EM is notoriously slow.
    \end{itquote}
    We included a comparison to other methods in terms of computation time on 
    one of the applications in the discussion. The comparison shows that our 
    method is competitive and outperforms some commonly used methods in terms of
    computation time.

    \begin{itquote}
      The authors have also missed several relevant references and 
      should put some more focus on the literature review.
    \end{itquote}
    We have included one of the suggested references in the introduction. 
    However, we feel that the other suggested references are not relevant and
    have thus not included them.

    \textbf{Reply to reviewer 2}
    \begin{itquote}
      This paper considered the use of group elastic net method. 
      Compared to Meier et al (2008, JRSS B) who used group lasso penalty, you 
      proposed to use the group elastic net. This extension is fine, but the 
      novelty is rather little in this sense.
    \end{itquote}
    The group lasso is built on an eniterely different philosophy.
    Estimation is very different and not adaptive. Group lasso was designed for
    small interpretable groups of features, such as dummy variables. Our method
    is designed for larger groups of features, where the grouping is based on 
    external information that may informative for the prediction problem at 
    hand. The group lasso has one penalty parameter for all groups and estimates
    it by cross-validation. Our method uses a different penalty parameter per 
    group of features and estimates them by empirical Bayes.

    \begin{itquote}
      In (2.2), the penalization parameters $\lambda_1$ , $\lambda_2$ 
      and $\lambda_g$ appears in the forms of $\lambda_1 \sqrt{\lambda_g'}$ and 
      $\lambda_2 \lambda_g'$. A question is, is there is an identifiability 
      issue here? If not, why not?
    \end{itquote}
    There is not, since we fix $\lambda_1$ and $\lambda_2$ beforehand through 
    cross-validation. This leaves $G$ penalties to estimate for $G$ groups with
    empirical Bayes, where each group consists of many features.

    \begin{itquote}
      The computational load and time: As commented in the Discussion,
      the proposed method uses the double EM loop, which not only increases the
      chance of ending up in a local optimum but also can be very time consuming
      in computation. It would be helpful by commenting on the computation time,
      for example, in the real data analysis.
    \end{itquote}
    We included comparisons in terms of computation time in the discussion that 
    show that our method is competitive in terms of computation time. In 
    addition, we investigated different starting values for our algorithm in the
    applications presented in the manuscript and supplementary material. In none
    of the applications did this lead to different optimisation results.

    \closing{Yours Faithfully,}
    \encl{manuscript.pdf, supplement.pdf}

  \end{letter}
\end{document}

### 2019-02-27 docs/supplement.Rnw
changed the applications in manuscript.Rnw and supplement.Rnw

  \subsection{Expectation and variance}
  The expectation and variance of $\beta$ are:
  \begin{align*}
  \E[\beta] &= g(\lambda_1,\lambda_2) \int_{-\infty}^{\infty} \beta \cdot \exp\
  -\frac{1}{2} (\lambda_1 |\beta| + \lambda_2 \beta^2)\] d\beta=0, \\
  \V[\beta] &= g(\lambda_1,\lambda_2) \int_{-\infty}^{\infty} \beta^2 \cdot
  \exp\[-\frac{1}{2} (\lambda_1 |\beta| + \lambda_2 \beta^2)\] d\beta -
  \E[\beta]^2\\ &= \frac{\lambda_1^2}{4 \lambda_2^2} + \frac{1}{\lambda_2} -
  \frac{\lambda_1}{2 \lambda_2^{3/2}} \phi \( \frac{\lambda_1}{2 
  \sqrt{\lambda_2}} \) \Phi \(\frac{-\lambda_1}{2 \sqrt{\lambda_2}} \)^{-1}.
  \end{align*}

@@

\section{Application to microRNAs in colorectal cancer}\label{sec:appcolon}
\subsection{Partitioning based on differential expression}\label{sec:mirna}
The data set is from a deep sequencing analysis on microRNA \cite[]{neerincx_combination_2018}. The study was done in 88 treated colorectal cancer patients, with the aim of classifying treatment response, coded as either non-progressive/remission (70 patients) or progressive (18 patients). After pre-processing and normalisation, 2114 microRNAs remained. In addition to the 2114 microRNAs we incorporated 4 unpenalized clinical covariates into the analysis: prior use of adjuvant therapy (binary), the type of systemic treatment regimen (ternary), age, and primary tumor differentiation (binary).

In a preliminary experiment on different subjects, the microRNA expression levels of metastatic colorectal tumour tissue were compared to normal non-colorectal tissue and primary colorectal tumour tissue was compared to primary colorectal tumour tissue \cite[]{neerincx_mir_2015}. This yielded 221 microRNAs that were differentially expressed in both comparisons ($\text{FDR} \leq 0.05$), versus 1893 not differentially expressed microRNAs. We expect that incorporation of this partitioning enhances therapy response classification, because tumor-specific miRNAs might be more relevant than non-specific ones. In addition, we divided the differentially expressed microRNAs even further into 127 highly differentially expressed microRNAs ($\text{FDR} \leq 0.001$) and 94 medium differentially expressed microRNAs ($0.001 < \text{FDR} \leq 0.05$).  We will refer to this second partitioning as the three-group setting, as opposed to the first, the two-group setting.

We compared \texttt{gren} to ridge and elastic net regression, and \texttt{GRridge}. For the elastic net methods (including \texttt{gren}) we set $\alpha \in \{ 0.05, 0.5, 0.95\}$. The estimated penalty multipliers are according to expectation: in all group-regularized methods, the 221 differentially expressed microRNAs receive the smallest penalty (Figure \ref{fig:col_bar}a). In the three-group setting the pattern is again as expected (Figure \ref{fig:col_bar}b): the highly expressed group, receives the lowest penalty, followed by the medium expressed group, while the non-expressed group receives the strongest penalty. \texttt{GRridge} is not able to distinguish between the medium and non-differentially expressed groups of microRNAs.

\begin{figure}[!h]
	\centering
	\includegraphics[width=0.83\linewidth]{grEBEN_mirseq_Maarten_bar1.png}
	\caption{Estimated penalty multipliers for (a) the two-group setting and (b) the three-group setting.}
	\label{fig:col_bar}
\end{figure}
Predictive performance of the methods is measured by AUC and BSS, both estimated by leave-one-out cross-validation (LOOCV), and shown in Figure \ref{fig:col_per}. All group-regularized elastic net models consistently outperform their non-group-regularized counterparts in both settings, in terms of AUC and BSS. In both settings, \texttt{gren} with $\alpha=0.5$ and $\alpha=0.95$ outperforms the other methods for a large range of model sizes in terms of AUC. This is especially true for the larger models. \texttt{gren} with $\alpha=0.5$ outperforms the other methods with respect to BSS in the smaller model ranges, while \texttt{gren} with $\alpha=0.05$ becomes competitive for the larger models. A general pattern in both settings is that \texttt{GRridge} and \texttt{gren} with $\alpha=0.05$ perform similarly. This is not surprising, because an elastic net model with $\alpha=0.05$ is close to a ridge model, in terms of penalisation. In general, the classifiers with three group penalties are better than the ones with two group penalties when relatively few features are used. Comparing, for example, the models of size 20 for \texttt{gren} with $\alpha=0.5$, we have an AUC of 0.68 for the two-group setting, versus an AUC of 0.78 for the three-group setting. Note that models with few features are often desirable for clinical implementation.
\begin{figure}[!h]
	\centering
	\includegraphics[width=1\linewidth]{grEBEN_mirseq_Maarten_combined_performance.png}
	\caption{Cross-validated performance as (a) AUC in the two-group setting, (b) Brier skill score in the two-group setting, (c) AUC in the three-group setting, and (d) Brier skill score in the three-group setting against the number of selected features for each method.}
	\label{fig:col_per}
\end{figure}
\begin{figure}[!h]
	\centering
	\includegraphics[width=0.83\linewidth]{grEBEN_mirseq_Maarten_bar1.png}
	\caption{Estimated penalty multipliers for (a) the two-group setting and (b) the three-group setting.}
	\label{fig:col_bar}
\end{figure}

Predictive performance of the methods is measured by AUC and BSS, both estimated by leave-one-out cross-validation (LOOCV), and shown in Figure \ref{fig:col_per}. All group-regularized elastic net models consistently outperform their non-group-regularized counterparts in both settings, in terms of AUC and BSS. In both settings, \texttt{gren} with $\alpha=0.5$ and $\alpha=0.95$ outperforms the other methods for a large range of model sizes in terms of AUC. This is especially true for the larger models. \texttt{gren} with $\alpha=0.5$ outperforms the other methods with respect to BSS in the smaller model ranges, while \texttt{gren} with $\alpha=0.05$ becomes competitive for the larger models. A general pattern in both settings is that \texttt{GRridge} and \texttt{gren} with $\alpha=0.05$ perform similarly. This is not surprising, because an elastic net model with $\alpha=0.05$ is close to a ridge model, in terms of penalisation. In general, the classifiers with three group penalties are better than the ones with two group penalties when relatively few features are used. Comparing, for example, the models of size 20 for \texttt{gren} with $\alpha=0.5$, we have an AUC of 0.68 for the two-group setting, versus an AUC of 0.78 for the three-group setting. Note that models with few features are often desirable for clinical implementation.
\begin{figure}[!h]
	\centering
	\includegraphics[width=1\linewidth]{grEBEN_mirseq_Maarten_combined_performance.png}
	\caption{Cross-validated performance as (a) AUC in the two-group setting, (b) Brier skill score in the two-group setting, (c) AUC in the three-group setting, and (d) Brier skill score in the three-group setting against the number of selected features for each method.}
	\label{fig:col_per}
\end{figure}

\section{Application to RNAseq in lymph node cancer}\label{sec:applymph}

CoRF data:

<<lines_rnaseq_lymph_node_metastasis_res1_auc, fig.cap="Cross-validated AUC in lymph node metastasis example, with a) the full range and b) cut-out", out.width="100%", fig.asp=2/3>>=
@

\subsection{Random groups}
Considering that the features may be partitioned into many groups, with one parameter per group, we have to be aware of overfitting risks. We investigated this using the data introduced in Section \ref{sec:mirna}, randomly dividing the features into three groups. We fixed the group sizes to the group sizes used in the three-group setting in Section \ref{sec:mirna}. Under this random partitioning of the features, we expect all penalty multipliers to be estimated as one if overfitting does not occur.

We compared the estimated multipliers to the estimates by \texttt{GRridge}. Since these results depend on one specific randomisation of the groups, we repeated the procedure 100 times and present the results in Figure \ref{fig:random1}. From this figure we see that the estimates for \texttt{gren} are close to one. The estimates by \texttt{GRridge} show much more variation. Additionally, \texttt{GRridge} gives slightly biased estimates: the median penalty parameter estimates are 1.16, 1, and 0.85 for the three groups. In contrast, the median estimates of \texttt{gren} are 1, 1, and 0.99. 
\begin{figure}[!h]
	\centering
	\includegraphics[width=0.5\linewidth]{grEBEN_mirseq_Maarten_res6_boxplot.png}
	\caption{Estimated penalty multipliers of the group-regularized methods for 100 repeats of 3 random groups in colorectal miRNA data.}
	\label{fig:random1}
\end{figure}

We repeated the simulation with ten evenly sized groups. Figure \ref{fig:random2} presents the estimated penalty multipliers for 100 random splits of the microRNA features into 10 evenly sized groups for \texttt{GRridge} and \texttt{gren} with $\alpha \in \{ 0.05, 0.5, 0.95 \}$. From the Figure we see that \texttt{GRridge} shows more variation around one then the other models. In addition, the median penalty multipliers for the three \texttt{gren} models are all between 0.99 and 1.02, while the \texttt{GRridge} median estimates range from 1.04 to 1.09. This indicates that \texttt{GRridge} suffers from a slight bias in the estimation, while the \texttt{gren} estimates appear to be unbiased.
\begin{figure}[!h]
	\centering
	\includegraphics[width=0.9\linewidth]{grEBEN_mirseq_Maarten_res8_boxplot.png}
	\caption{Estimated penalty multipliers of the group-regularized methods for 100 repeats of 10 random groups in colorectal miRNA data.}
	\label{fig:random2}
\end{figure}

\subsection{Stability of selection}
One common criticism on the elastic net, and especially the lasso, is the instability of feature selection. That is, different data sets yield different sets of selected features. To investigate the stability of feature selection by \texttt{gren} we created 50 stratified bootstrap samples from the microRNA data \cite[]{neerincx_combination_2018}, where we stratified by treatment response. For each of these bootstrap samples we estimated the regular elastic net and \texttt{gren} for $\alpha \in \{0.05, 0.5, 0.95 \}$, where we selected 25 features (including the 5 unpenalized covariates). We calculated the size of all ${50}\choose{2}$ intersections of selected features and present the results in Figure \ref{fig:bootstrap}. 
\begin{figure}[!h]
	\centering
	\includegraphics[width=0.9\linewidth]{gren_mirseq_Maarten_overlap.png}
	\caption{Overlap between stratified bootstrap samples for the elastic net models with (a) $\alpha=0.05$, (b) $\alpha=0.5$, and (c) $\alpha=0.95$.}
	\label{fig:bootstrap}
\end{figure}
From this figure it can be seen that stability of feature selection in \texttt{gren} is higher than in the regular elastic net for all three $\alpha$'s.

>>>>

  \subsection{Expectation and variance}
  The expectation and variance of $\beta$ are:
  \begin{align*}
  \E[\beta] &= g(\lambda_1,\lambda_2) \int_{-\infty}^{\infty} \beta \cdot \exp\[
  -\frac{1}{2} (\lambda_1 |\beta| + \lambda_2 \beta^2)\] d\beta=0, \\
  \V[\beta] &= g(\lambda_1,\lambda_2) \int_{-\infty}^{\infty} \beta^2 \cdot
  \exp\[-\frac{1}{2} (\lambda_1 |\beta| + \lambda_2 \beta^2)\] d\beta -
  \E[\beta]^2\\ &= \frac{\lambda_1^2}{4 \lambda_2^2} + \frac{1}{\lambda_2} -
  \frac{\lambda_1}{2 \lambda_2^{3/2}} \phi \( \frac{\lambda_1}{2 
  \sqrt{\lambda_2}} \) \Phi \(\frac{-\lambda_1}{2 \sqrt{\lambda_2}} \)^{-1}.
  \end{align*}

@@

  \section{Application to MicroRNAs in colorectal cancer}\label{sec:colorectal}
  In this section we present additional results to Section [! section ref here]
  in the Main Document.

  \subsection{Random groups}
  Considering that the features may be partitioned into many groups, with one 
  parameter per group, we have to be aware of overfitting risks. We investigated
  this using the data introduced in Section [! section ref here] of the Main 
  Document, randomly dividing 
  the features into three groups. We fixed the group sizes to the group sizes 
  used in the three-group setting in Section [! section ref here] of the Main
  Docuemtn. Under this random 
  partitioning of the features, we expect all penalty multipliers to be 
  estimated as one if no overfitting occurs.

  We compared the estimated multipliers to the estimates by \texttt{GRridge}. 
  Since these results depend on one specific randomisation of the groups, we 
  repeated the procedure 100 times and present the results in [! figure ref]. 
  From this figure we see that the estimates for \texttt{gren} are close to one.
  The estimates by \texttt{GRridge} show much more variation. Additionally, 
  \texttt{GRridge} gives slightly biased estimates: the median penalty parameter
  estimates are [! medians here] for the three groups. In contrast, the median 
  estimates of \texttt{gren} are [! medians here].

  [! figure here]

  We repeated the simulation with ten evenly sized groups. Figure 
  [! figure ref here] presents the estimated penalty multipliers for 100 random 
  splits of the microRNA features into 10 evenly sized groups for 
  \texttt{GRridge} and \texttt{gren} with $\alpha \in \{ 0.05, 0.5, 0.95 \}$. 
  From the Figure we see that \texttt{GRridge} shows more variation around one 
  then the other models. In addition, the median penalty multipliers for the 
  three \texttt{gren} models are all between [! medians here], while the 
  \texttt{GRridge} median estimates range from [! medians here]. This indicates 
  that \texttt{GRridge} suffers from a slight bias in the estimation, while the 
  \texttt{gren} estimates appear to be unbiased.

  [! figure here]

  \subsection{Stability of selection}
  One common criticism on the elastic net, and especially the lasso, is the 
  instability of feature selection. That is, different data sets yield different
  sets of selected features. To investigate the stability of feature selection 
  by \texttt{gren} we created 50 stratified bootstrap samples from the microRNA 
  data \cite[]{neerincx_combination_2018}, where we stratified by the target of 
  prediction, treatment response. 
  For each of these bootstrap samples we estimated the regular elastic
  net and \texttt{gren} for $\alpha \in \{0.05, 0.5, 0.95 \}$, where we selected
  25 features (including the 5 unpenalized covariates). We calculated the size 
  of all ${50}\choose{2}$ intersections of selected features and present the 
  results in Figure [! figure ref here].

  [! figure here]

  \section{Application to metabolomics in alzheimer}\label{sec:alzheimer}
	We applied our method to an Alzheimer's dataset from 
	\cite{de_leeuw_blood-based_2017}. The
	data consist of 87 diagnosed Alzheimer's patients with the $\epsilon$4
	allele of the APOE gene and 87 subjects with subjective un-diagnosed cognitive
	decline, without the $\epsilon$4 allele. The aim of the study was to diagnose 
	subjects with Alzheimer's disease based on their metabolic profile. After 
	pre-processing, 230 metabolite expression levels were obtained. To enhance
	classification performance, we included two sources of co-data: (a) the 
	relative standard devitation of the measurements (RSD), a commonly 
	used quality score (smaller is better) for metabolites, binned into five 
	groups, and (b) the node degree of the metabolites, based on a differential 
	network analysis of the metabolites, binned into three categories: 0 degree, 
	positive, lower than average degree, and postive, higher than average degree. 
	We expect a lower RSD and a higher node degree to be indicative of an 
	important metabolite and therefore receive less penalty.

	To estimate performance measures AUC and BSS, we split the data into 120 
	training instances and 54 test instances. 
	The training data is used to fit the models, while the test data
	is used to estimate the performance measures on. We included the different 
	co-data types seperately and combined, which resulted in the estimated 
	multipliers and performance measures presented in Figure 
	\ref{fig:barplot_lines_metabolomics_alzheimer_res5_auc}.
<<barplot_lines_metabolomics_alzheimer_res5_auc, fig.cap="Results for the alzheimer example with (a) the estimated penalty multipliers for the quality score, (b) the estimated penalty multipliers for the network degree, (c) the estimated AUC, and (d) the estimated Brier skill score.", out.width="100%", fig.asp=1>>=
@

  In this example, \texttt{GRridge} and \texttt{gren} underperform due to a 
  small number of features to learn the penalty parameters from. The other 
  methods estimate one global penalty parameter, so do not suffer from the 
  lower dimensionality of the problem.

  \section{Application to microRNAs in cervical cancer}
  A deep sequencing analysis on small non-coding ribonucleic acid (microRNAseq) 
  was performed on 24 women with high-grade cervical intraepithelial 
  neoplasia (CIN3) and 32 healthy women for the purpose of finding relevant 
  screening markers for cervical cancer screening. The next generation 
  sequencing analysis resulted in 2,576 transcripts. The data was normalized 
  and pre-processed, rendering 752 transcripts. More details of the
  data sets and the pre-processing steps are found in the supplementary   
  material of \cite{novianti_better_2017}.

  The 752 microRNAs were divided in three classes based on their conservation 
  status. Conservation status refers to whether the microRNAs are only found in
  humans or are conserved across different species. They come in three classes:
  (a) non-conserved (535), (b) conserved in mammals (70), and (c) 
  conserved across most vertebrates (147). We expect (c) to contain the most 
  important microRNAs, followed by (b) and (a), because these microRNAs have not
  been lost during evolution and might therefore regulate important processes.

<<barplots_micrornaseq_cervical_cancer_res1_auc, fig.cap="Estimated penalty multipliers for conservation status co-data", out.width="50%", fig.asp=2/3>>=
@

<<lines_micrornaseq_cervical_cancer_res1_auc, fig.cap="Cross-validated AUC in cervicl cancer example", out.width="50%", fig.asp=1>>=
@
